Physcraper: A Python package for continually updated phylogenetic trees using the Open Tree of Life

### 2021-01-07 sources/bmc_template_submission/bmc_article.tex
update intro

%%%%%%%%%%%%%%%%
%% Background %%
%%
\section*{Background}
Phylogenies capture the shared history of organisms and provide key evolutionary
context for our biological observations \cite{dobzhansky1973nothing}. Data such as
geographical location, fossil ranges, and genetic and phenotypic information increasingly
available in public databases constitute an amazing
resource for biological discovery \cite{baxevanis2015importance}.
Connecting existing phylogenies with public biological data in a reproducible and
continuous manner would represent a key tool for evolutionary studies.

Here, we introduce Physcraper, a tool that automates biological database connections
with the main goal of building upon homology
hypotheses that taxon specialists have assessed and deemed appropriate for a
specific phylogenetic scope to update a starting tree and single locus alignments
with public DNA data, with the added capability to connect the lineages represented
in the phylogeny to biological data from public databases.

One of the main challenges for automatic
integration of biological data into phylogenies are varying taxonomic idiosyncrasies
across databases. To address this challenge, the Open Tree of Life project (OpenTree)
created a unified taxonomy for name standardization, by integrating taxonomic
data from several databases \cite{rees2017automated}, including the USA National
Center for Biodiversity Information (NCBI) molecular database GenBank
\cite{benson2000genbank}, among others.
Existing OpenTree taxonomy programmatic tools
are a key resource that can be used to establish a framework for connecting data from any
biological database that has been integrated to OpenTree's unified taxonomy.

OpenTree's unified taxonomy is used to construct OpenTree's comprehensive tree of
life that synthesizes published phylogenies and taxonomy. It comprises 2.3 million tips, of which
around 90,000 (87,740) are supported by phylogenies - the remaining 1.4 million taxa are
placed in the tree based on taxonomy.
This large gap in public phylogenetic knowledge is due to several factors:
- lack of data sharing, most analyses essentially lost forever;
- lack of data, we still lack genetic data for a large portion of organisms; and
- lack of analyticla power of existing data.

Decades of single locus sequencing have generated
massive amounts of homologous DNA datasets that have the potential to be used for phylogenetic
reconstruction at many scales \cite{sanderson2008phylota}.
More than a decade ago, GenBank release 159 (April 15, 2007) already hosted 72
million DNA sequences that were gauged to have the potential to resolve phylogenetic
relationships of 98.05\% of the almost 241,000
distinct taxa in the NCBI taxonomy at the time \cite{sanderson2008phylota}.

With more than 221 million single locus
sequences in GenBank today \cite{genbankstats}, there is a considerable amount of phylogenetically
informative data in GenBank that has not been analysed and/or made publicly available
with the potential to significantly fill phylogenetic gaps in the tree of life.

Assembling a DNA alignment from such a massive database can be done "by hand",
but it is a largely time consuming and mostly non-reproducible approach.
Computational pipelines that mine DNA databases fast, efficiently, and reproducibly
have been applied to infer phylogenetic relationships in a variety of organisms
(e.g., \cite{smith2009mega, antonelli2017toward, izquierdo2014pumper}).
While genomics has, and will continue to, revolutionize phylogenetic inference,
the diversity of alternative genomic sequencing approaches implemented produce
largely non-overlapping homology hypotheses across taxa \cite{jones2016targeted},
creating challenges for phylogenetic reconstruction.
Phylogenomics addresses this problem by focusing on targeted capture of informative
regions \cite{andermann2020guide}.
However, fine-grained curated markers and alignments can significantly improve phylogenetic
reconstructions, even in phylogenomic analyses \cite{fragoso2017pilot}.

There are almost 8,200 publicly available, peer-reviewed curated alignments, covering
around 100,000 distinct taxa in the TreeBASE database \cite{piel2009treebase}, which
can be leveraged as seeds to mine molecular databases, and as "jump-start" alignments
for phylogenetic reconstructions \cite{morrison2006multiple} to continually enrich,
update and compare existing phylogenetic knowledge.

Such an approach was proposed in PUMPER but the community has not picked it up,
probably due to installation difficulties.
Physcraper extends this approach by allowing for interoperability.

Physcraper is implemented as a Python pipeline using OpenTree's taxonomy and programmatic access
protocols (API's) to implement a database interoperability framework that automatically
links phylogenies that have been standardized to OpenTree taxonomy, to alignments
from TreeBASE, data from GenBank, and phylogenies from OpenTree's Phylesystem.
Physcraper aims to demonstrate the benefits of reproducible workflows and open
science in phylogenetics, and encourage better data sharing practices in the community.

>>>>

%%%%%%%%%%%%%%%%
%% Background %%
%%
\section*{Background}
Phylogenies capture the shared history of organisms and provide key evolutionary
context for our biological observations \cite{dobzhansky1973nothing}.
Updating existing phylogenies with with molecular sequence data that has never been 
incorporated into any phylogenetic estimate provides the opportunity to capture the evolutionary history of many taxa.
Here, we introduce Physcraper, a tool that automates database connections
to build on published alignments to extend phylogenetic inferences with more taxa.
Physcraper can be used to to update a starting tree and single locus alignments
with public DNA data, and link the tips in these trees to a unified, interoperable taxonomic resource \cite{rees2017automated}.

Data such as
geographical location, fossil ranges, and genetic and phenotypic information increasingly
available in public databases which constitute an amazing
resource for biological discovery \cite{baxevanis2015importance}.
One of the main challenges for automatic
integration of biological data across databases are varying taxonomic idiosyncrasies. 
To address this challenge, the Open Tree of Life project (OpenTree)
created a unified taxonomy for name standardization, by integrating taxonomic
data from several databases \cite{rees2017automated}, including the USA National
Center for Biodiversity Information (NCBI) taxonomy \cite{federhen_ncbi_2012, schoch_ncbi_2020}, and the Global Biodiversity Information Facility (GBIF) \cite{gbif_secretariat_gbif_2019} among others.
By using the existing OpenTree taxonomy programmatic tools to map tip names,
Physcraper has a framework for connecting updated phylogenies data from any
biological database.

Decades of single locus sequencing have generated
massive amounts of homologous DNA datasets that have the potential to be used for phylogenetic
reconstruction at many scales \cite{sanderson2008phylota}.
More than a decade ago, GenBank release 159 (April 15, 2007) already hosted 72
million DNA sequences that were gauged to have the potential to resolve phylogenetic
relationships of 98.05\% of the almost 241,000
distinct taxa in the NCBI taxonomy at the time \cite{sanderson2008phylota}.
However, even thirteen years later, phylogenetic estimates for many of these taxa are not available \cite{mctavish_how_2017}.
OpenTree's comprehensive tree of life comprises 2.3 million tips, of which around 90,000 are supported by phylogenies - the remaining 1.4 million taxa are placed in the tree based on taxonomy.
There is a considerable amount of phylogenetically
informative data in GenBank with the potential to fill these phylogenetic gaps in the tree of life that, but this data either has not been analysed or the analyses have not been made publicly available \cite{mctavish_how_2017}.

Assembling a DNA alignment from such a massive database can be done "by hand",
but that is a time consuming approach which is not highly reproducible.
A variety of computational pipelines that mine DNA databases fast, efficiently, and reproducibly
have been applied to infer phylogenetic relationships in a variety of organisms
(e.g., \cite{smith2009mega, antonelli2017toward, izquierdo2014pumper}).
While genomics has, and will continue to, revolutionize phylogenetic inference,
the diversity of alternative genomic sequencing approaches implemented produce
largely non-overlapping homology hypotheses across taxa \cite{jones2016targeted},
creating challenges for phylogenetic reconstruction.
Phylogenomics addresses this problem by focusing on targeted capture of informative
regions \cite{andermann2020guide}.
However, fine-grained curated markers and alignments can significantly improve phylogenetic
reconstructions, even in phylogenomic analyses \cite{fragoso2017pilot}.

Physcraper improves on previous work in automating phylogenetics by leveraging the power of existing homology hypotheses that taxon specialists have assessed and deemed appropriate for a specific phylogenetic scope.
There are almost 8,200 publicly available, peer-reviewed curated alignments, covering
around 100,000 distinct taxa in the TreeBASE database \cite{piel2009treebase}, which
can be leveraged as seeds to mine molecular databases, and as ``jump-start'' alignments
for phylogenetic reconstructions \cite{morrison2006multiple} to continually enrich,
update and compare existing phylogenetic knowledge.

Physcraper is a Python pipeline using OpenTree's programmatic access
protocols (API's) to automatically
links phylogenies mapped to a standardized taxonomy \cite{mctavish2015phylesystem}, to alignments
from TreeBASE \cite{sanderson1994treebase}, data from GenBank \cite{benson2000genbank}.
Physcraper provides automated and reproducible workflows for updating existing phylogenetic estimates.
